---
title: 'RR: Assessing Health and Economic Impact Of Weather Events'
author: "Olagunju Ayodeji"
date: "February 3, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis of Study Results

This study U.S. National Oceanic and Atmospheric Administration's
(NOAA) storm database. This database tracks characteristics of major
storms and weather events in the United States, including when and
where they occur, as well as estimates of any fatalities, injuries,
and property damage.

Wind events, which include tornadoes and hurricanes, are by the far
the most harmful in aggregate, causing over 100,000 injuries and 90
deaths over the course of this study.  Though less frequent, severe
heat events have the highest incidence of deaths and injuries per
event.  This study finds that severe rain and wind events are by far
the most costly in terms of dollars spent to replace property and crop
damage.

## Questions this study considers

  1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
  2. Across the United States, which types of events have the greatest economic consequences?

## Notes about the environment used
This study was done using the following tools, including OS and Programming language versions

The study was conducted on a 32-bit Windows 7 machine with 2 cores.

R language was `r R.version.string`

For publishing to `rpubs.com`, I used RStudio version 1.1.463

The full project may be found on Github at `https://github.com/djolas/05-reproducible-research-assignment-2`

## Data Processing

There will be categorization of event type that may be subjective.   Not to mention the data collection
of the weather events will be categorized on the field manually and hence subjected to human input errors.

```{r Set my working Dir}
setwd ("C:/Users/djolas/My Documents/data")
```

```{r Downloading Data from Internet}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,destfile,mode="wb")
url2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
destfile2 <- "repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
download.file(url2,destfile2,mode="wb")
```

```{r Read the file in BZ format - cool! - and then check basic EDA}
StormData<- read.csv(bzfile(destfile), stringsAsFactors = FALSE)
names(StormData)
summary(StormData$FATALITIES)
summary(StormData$INJURIES)
```

It seems that there are 985 unique types of events, I will endeavor to decrease this list a lot more
I will also create a list for Question 1 for most harmful and Question 2 for damages PROPDMGEXP,CROPDMGEXP
Also, I will only need certain columns FATALITIES INJURIES PROPDMG CROPDMG PROPDMGEXP CROPDMGEXP

``` {r Keeping on the needed columns and Row using SUBSET}
KeepCol <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG","CROPDMG","PROPDMGEXP","CROPDMGEXP")
StormDataKeep <- subset(StormData, select = KeepCol)
names(StormDataKeep)
StormDataKeepNZ <- subset(StormDataKeep, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)
```


```{r New Rollup the EVTYPE - Fix the actual numbers as well}
StormDataKeepNZ$EVENT <- StormDataKeepNZ$EVTYPE
StormDataKeepNZ$EVENT[grep("flood",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("warm",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("freeze",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("avala",StormDataKeepNZ$EVENT, ignore.case = T)] <-"AVALANCHE"
StormDataKeepNZ$EVENT[grep("ice",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("bitter",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("BELOW NORMAL PRECIPITATION",StormDataKeepNZ$EVENT, ignore.case = T)] <-"DROUGHT"
StormDataKeepNZ$EVENT[grep("BLIZZARD",StormDataKeepNZ$EVENT, ignore.case = T)] <-"SNOW"
StormDataKeepNZ$EVENT[grep("SNOW",StormDataKeepNZ$EVENT, ignore.case = T)] <-"SNOW"
StormDataKeepNZ$EVENT[grep("dry",StormDataKeepNZ$EVENT, ignore.case = T)] <-"DROUGHT"
StormDataKeepNZ$EVENT[grep("BEACH",StormDataKeepNZ$EVENT, ignore.case = T)] <-"BEACH EROSION"
StormDataKeepNZ$EVENT[grep("FIRE",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FIRE"
StormDataKeepNZ$EVENT[grep("COLD",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("RAIN",StormDataKeepNZ$EVENT, ignore.case = T)] <-"RAIN"
StormDataKeepNZ$EVENT[grep("CHILL",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("wet",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("frost",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("freez",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("gust",StormDataKeepNZ$EVENT, ignore.case = T)] <-"WIND"
StormDataKeepNZ$EVENT[grep("hail",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HAIL"
StormDataKeepNZ$EVENT[grep("heat",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("hurricane",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HURRICANE"
StormDataKeepNZ$EVENT[grep("typhoon",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HURRICANE"
StormDataKeepNZ$EVENT[grep("ice",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("icy",StormDataKeepNZ$EVENT, ignore.case = T)] <-"COLD"
StormDataKeepNZ$EVENT[grep("LANDSLIDE",StormDataKeepNZ$EVENT, ignore.case = T)] <-"LANDSLIDES"
StormDataKeepNZ$EVENT[grep("mud",StormDataKeepNZ$EVENT, ignore.case = T)] <-"MUDSLIDES"
StormDataKeepNZ$EVENT[grep("high temp",StormDataKeepNZ$EVENT, ignore.case = T)] <-"HEAT"
StormDataKeepNZ$EVENT[grep("TROPICAL STORM",StormDataKeepNZ$EVENT, ignore.case = T)] <-"TROPICAL STORM"
StormDataKeepNZ$EVENT[grep("light",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("tstm",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("torn",StormDataKeepNZ$EVENT, ignore.case = T)] <-"TORNADO"
StormDataKeepNZ$EVENT[grep("tide",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("tsu",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$EVENT[grep("thun",StormDataKeepNZ$EVENT, ignore.case = T)] <-"THUNDERSTORM"
StormDataKeepNZ$EVENT[grep("tsu",StormDataKeepNZ$EVENT, ignore.case = T)] <-"FLOOD"
StormDataKeepNZ$PropertyDamageAmt <- StormDataKeepNZ$PROPDMG
MultLookup <- c(M = 10^6, m = 10^6, K = 10^3, k = 10^3, B = 10^9, b = 10^9)
StormDataKeepNZ$PropertyDamageAmt <- StormDataKeepNZ$PROPDMG * MultLookup [as.character(StormDataKeepNZ$PROPDMGEXP)]
StormDataKeepNZ$CropDamageAmt     <- StormDataKeepNZ$CROPDMG * MultLookup [as.character(StormDataKeepNZ$CROPDMGEXP)]
StormDataKeepNZ$TotDamageAmt   <- StormDataKeepNZ$PropertyDamageAmt + StormDataKeepNZ$CropDamageAmt
```

====================================================================

# RESULTS
Now we will aggregate the results and output the top 10 events graphically.  This will answer the
2 questions

``` {r Aggregate and Graph - somehow the aggregate function does not work - using plyr - get top 10}
library(plyr)
SFatal <- ddply(StormDataKeepNZ, .(EVENT), summarize, DEATHS = sum(FATALITIES, na.rm = TRUE))
SInjury <- ddply(StormDataKeepNZ, .(EVENT), summarize, INJURED = sum(INJURIES, na.rm = TRUE))
SPropDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, PROPERTYCOST = sum(PropertyDamageAmt, na.rm = TRUE))
SCropDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, CROPCOST = sum(CropDamageAmt, na.rm = TRUE))
STotDam <- ddply(StormDataKeepNZ, .(EVENT), summarize, TOTALCOST = sum(TotDamageAmt, na.rm = TRUE))
SDEATH10 <- head(SFatal[order(-SFatal$DEATHS),],10)
SINJURED10 <- head(SInjury [order(-SInjury$INJURED),],10)
SPropDam10 <- head(SPropDam[order(-SPropDam$PROPERTYCOST),],10)
SCropDam10 <- head(SCropDam[order(-SCropDam$CROPCOST),],10)
STotDam10 <- head(STotDam[order(-STotDam$TOTALCOST),],10)
SDEATH10
SINJURED10
SPropDam10
SCropDam10
STotDam10
```
=========================================================

## Graphing the results

```{r results1,fig.height=20,fig.width=20,cache=TRUE}
library (ggplot2)
ggplot(SDEATH10,aes(EVENT)) + 
	geom_bar(aes(weight=DEATHS, fill =..count..)) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("FATALITIES") + 
	ggtitle("FATALITIES BY EVENTS")
```

```{r results2,fig.height=20,fig.width=20,cache=TRUE}
ggplot (SINJURED10,aes(EVENT)) + 
	geom_bar(aes(weight=INJURED, fill =..count..),) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("INJURIES") + 
	ggtitle("INJURIES BY EVENTS")
```

```{r results3,fig.height=20,fig.width=20,cache=TRUE}
ggplot (STotDam10,aes(EVENT)) + 
	geom_bar(aes(weight=TOTALCOST, fill =..count..)) + 
	xlab("EVENTS") + theme(axis.text.x=element_text(angle = 90)) + 
	ylab("DAMAGE COST") + 
	ggtitle("TOTAL DAMAGE COST")
```

	
================================================================================

## SUMMARY

The main causes of property damage are floods, hurricanes and tornadoes
Crop damages are generally caused by drought, flood, and cold

Fatalities are far and away caused by tornadoes and thunderstorms
Injuries are also far and away caused by tornadoes, with heat, thunderstorms.